---
title: "TRACE FT-ICR Temporal Slopes and Autocorrelation Workbook"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

## tl;dr

This notebook is an interactive companion for the Q2 temporal-structure analysis.
It refits:

- Pairwise distance slopes versus `log2(days between samples)`.
- Scalar semivariance slopes versus `log2(days between samples)`.
- Treatment-depth CAR(1) temporal autocorrelation terms, including all van Krevelen class fractions.

Run chunks top-to-bottom, or edit the parameters in the setup chunk and rerun selected sections.

## Setup

```{r setup}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 6,
  dpi = 160
)

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(cowplot)
  library(scales)
  library(nlme)
})

# Edit this path if the project moves.
project_dir <- normalizePath("/Users/jrweverk/Documents/TRACE_MCMC", winslash = "/", mustWork = TRUE)
docs_dir <- normalizePath(file.path(project_dir, ".."), winslash = "/", mustWork = TRUE)
out_dir <- file.path(project_dir, "fticr_integration", "output")
fig_dir <- file.path(out_dir, "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

# Interactive toggles.
write_outputs <- TRUE
use_cached_autocorrelation <- FALSE

# Pick one row for the CAR(1) decay-curve section.
selected_metric <- "mean_nosc"
selected_treatment <- "Warmed"
selected_depth <- "10cm"

depth_levels <- c("0-10 cm", "10-30 cm")
depth_display_labels <- c("0-10 cm" = "10cm", "10-30 cm" = "30cm")
treatment_cols <- c("Control" = "#2B6CB0", "Warmed" = "#C45A2A")
depth_shapes <- c("0-10 cm" = 16, "10-30 cm" = 17)
depth_offsets <- c("0-10 cm" = 0.14, "10-30 cm" = -0.14)
```

## Context & Methods

The pairwise distance and semivariance models use:

```text
response ~ log2(days between samples)
```

The slope is therefore the expected increase in distance, or scaled semivariance, per doubling of the sample interval.

The continuous-time autocorrelation model uses an irregular-time CAR(1) structure:

```text
Corr(e_i, e_j) = phi_30d ^ (|date_i - date_j| / 30)
```

So the reported autocorrelation term is the fitted 30-day correlation, but it defines a full exponential decay curve over arbitrary lags.

## Helper Functions

```{r helpers}
safe_numeric <- function(x) suppressWarnings(as.numeric(x))

classify_vk <- function(hc, oc, ai_mod) {
  fifelse(
    ai_mod >= 0.67,
    "condensed_aromatic",
    fifelse(
      ai_mod >= 0.5,
      "aromatic",
      fifelse(
        hc >= 1.5 & oc <= 0.3,
        "lipid",
        fifelse(
          hc >= 1.5 & oc > 0.3,
          "carbohydrate",
          fifelse(
            hc < 1.5 & hc >= 0.7 & oc <= 0.67,
            "lignin_like",
            fifelse(hc < 0.7, "condensed_aromatic", "other")
          )
        )
      )
    )
  )
}

transform_metric <- function(metric, value_plot) {
  if (metric == "richness") return(log1p(value_plot))
  if (metric == "frac_n_containing" || grepl("^vk_frac_", metric)) {
    p <- value_plot / 100
    p <- pmin(pmax(p, 1e-4), 1 - 1e-4)
    return(qlogis(p))
  }
  value_plot
}

standardize <- function(x) {
  s <- sd(x, na.rm = TRUE)
  if (!is.finite(s) || s == 0) return(rep(NA_real_, length(x)))
  (x - mean(x, na.rm = TRUE)) / s
}

clean_metric_label <- function(x) {
  gsub("\\n\\([^)]*\\)", "", gsub("\\n", " ", x))
}

theme_trace <- function(base_size = 10.5) {
  theme_minimal(base_size = base_size) +
    theme(
      legend.position = "top",
      legend.justification = "left",
      legend.title = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_blank(),
      strip.text = element_text(face = "bold", color = "#202124"),
      axis.text.y = element_text(color = "#202124"),
      axis.title = element_text(color = "#202124")
    )
}

fit_lm_slope <- function(dt, response_col) {
  mod_dt <- dt[is.finite(get(response_col)) & is.finite(log2_lag) & lag_days > 0]
  if (nrow(mod_dt) < 10L || uniqueN(mod_dt$log2_lag) < 2L) {
    return(data.table(
      n_pairs = nrow(mod_dt),
      intercept = NA_real_,
      slope_per_log2_day_interval = NA_real_,
      r2 = NA_real_
    ))
  }
  fit <- lm(mod_dt[[response_col]] ~ mod_dt$log2_lag)
  data.table(
    n_pairs = nrow(mod_dt),
    intercept = unname(coef(fit)[[1]]),
    slope_per_log2_day_interval = unname(coef(fit)[[2]]),
    r2 = summary(fit)$r.squared
  )
}

fit_distance_slopes <- function(pair_dt, distance_cols, group_cols) {
  groups <- unique(pair_dt[, ..group_cols])
  out <- list()
  idx <- 1L
  for (g_i in seq_len(nrow(groups))) {
    group_row <- groups[g_i]
    sub <- pair_dt[group_row, on = group_cols, nomatch = 0]
    for (distance_col in distance_cols) {
      out[[idx]] <- cbind(
        group_row,
        data.table(distance_metric = distance_col),
        fit_lm_slope(sub, distance_col)
      )
      idx <- idx + 1L
    }
  }
  rbindlist(out, fill = TRUE)
}

fit_scalar_semivariance_slopes <- function(scalar_pairs) {
  scalar_pairs[, log2_lag := log2(lag_days)]
  out <- scalar_pairs[
    ,
    fit_lm_slope(.SD, "scaled_semivar"),
    by = .(metric, metric_label, metric_order, treatment_label, depth_label)
  ]
  setorder(out, metric_order, depth_label, treatment_label)
  out
}

fit_car1_by_treatment_depth <- function(long_dt, metric_order_dt) {
  ctrl <- lmeControl(opt = "optim", msMaxIter = 200, maxIter = 100, returnObject = TRUE)
  rows <- list()
  idx <- 1L

  for (metric_name in metric_order_dt$metric) {
    metric_dt <- long_dt[metric == metric_name]
    for (dep in depth_levels) {
      for (trt in c("Control", "Warmed")) {
        d <- copy(metric_dt[depth_label == dep & treatment_label == trt])
        if (nrow(d) < 10L) next
        d[, date := as.IDate(date)]
        d[, series_id := factor(interaction(plot, depth_label, drop = TRUE))]
        d[, time30 := as.numeric(date - min(date, na.rm = TRUE)) / 30]
        d[, z := standardize(transform_metric(metric_name, value_plot))]
        d <- d[is.finite(z)]
        if (nrow(d) < 10L || uniqueN(d$series_id) < 2L) next

        fit0 <- try(lme(
          z ~ 1,
          random = ~1 | series_id,
          data = d,
          method = "ML",
          control = ctrl
        ), silent = TRUE)
        fit1 <- try(lme(
          z ~ 1,
          random = ~1 | series_id,
          correlation = corCAR1(value = 0.5, form = ~time30 | series_id),
          data = d,
          method = "ML",
          control = ctrl
        ), silent = TRUE)

        phi <- if (!inherits(fit1, "try-error")) {
          as.numeric(coef(fit1$modelStruct$corStruct, unconstrained = FALSE))
        } else {
          NA_real_
        }

        rows[[idx]] <- data.table(
          metric = metric_name,
          metric_label = unique(d$metric_label)[[1]],
          metric_order = unique(d$metric_order)[[1]],
          metric_group = unique(d$metric_group)[[1]],
          treatment_label = trt,
          depth_label = dep,
          n_samples = nrow(d),
          n_series = uniqueN(d$series_id),
          phi_per_30d = phi,
          half_life_days = if (is.finite(phi) && phi > 0 && phi < 1) 30 * log(0.5) / log(phi) else NA_real_,
          days_to_corr05 = if (is.finite(phi) && phi > 0 && phi < 1) 30 * log(0.05) / log(phi) else NA_real_,
          delta_aic_car1_minus_null = if (!inherits(fit0, "try-error") && !inherits(fit1, "try-error")) AIC(fit1) - AIC(fit0) else NA_real_,
          fit_ok = !inherits(fit1, "try-error")
        )
        idx <- idx + 1L
      }
    }
  }

  rbindlist(rows, fill = TRUE)
}
```

## Data Sources

```{r paths}
pairwise_distance_pairs_path <- file.path(out_dir, "fticr_q2_pairwise_temporal_distance_pairs.csv")
scalar_pairs_path <- file.path(out_dir, "fticr_formula_count_autocorrelation_pairs.csv")
long_metric_path <- file.path(out_dir, "fticr_formula_count_temporal_metrics_plot_data.csv")
cached_autocorr_path <- file.path(out_dir, "fticr_q2_autocorrelation_terms_by_treatment_depth_vk_classes.csv")
cached_autocorr_metric_path <- file.path(out_dir, "fticr_q2_autocorrelation_metric_plot_data_vk_classes.csv")

mol_path <- file.path(
  docs_dir,
  "trace_data_cleanup",
  "data_raw",
  "processed_data",
  "Merged_Output",
  "Test_Processed-Unprocessed_Mol.csv"
)
intensity_path <- file.path(
  docs_dir,
  "trace_data_cleanup",
  "data_raw",
  "processed_data",
  "Merged_Output",
  "Test_Processed-Unprocssed_Data.csv"
)

required_paths <- c(pairwise_distance_pairs_path, scalar_pairs_path, long_metric_path, mol_path, intensity_path)
missing_paths <- required_paths[!file.exists(required_paths)]
if (length(missing_paths) > 0L) {
  stop("Missing required input files:\n", paste(missing_paths, collapse = "\n"))
}

data.table(
  input = basename(required_paths),
  path = required_paths,
  exists = file.exists(required_paths)
)
```

## Day-Interval Slopes

### Refit Distance Slopes

```{r fit-distance-slopes}
distance_pairs <- fread(pairwise_distance_pairs_path)
distance_cols <- c("formula_jaccard_distance", "vk_class_braycurtis", "metric_vector_distance")

distance_slope_td <- fit_distance_slopes(
  distance_pairs,
  distance_cols = distance_cols,
  group_cols = c("treatment_label", "depth_label")
)

distance_names <- c(
  formula_jaccard_distance = "Formula identity\nJaccard",
  vk_class_braycurtis = "Chemical class\nBray-Curtis",
  metric_vector_distance = "Scalar metric\nvector"
)
distance_levels <- c("Formula identity\nJaccard", "Chemical class\nBray-Curtis", "Scalar metric\nvector")
distance_slope_td[, distance_label := factor(distance_names[distance_metric], levels = rev(distance_levels))]
distance_slope_td[, treatment_label := factor(treatment_label, levels = c("Control", "Warmed"))]
distance_slope_td[, depth_label := factor(depth_label, levels = depth_levels)]

distance_slope_td[
  ,
  .(
    distance_metric,
    treatment_label,
    depth_label,
    n_pairs,
    intercept,
    slope_per_log2_day_interval,
    r2
  )
]
```

### Refit Scalar Semivariance Slopes

```{r fit-scalar-semivariance-slopes}
scalar_pairs <- fread(scalar_pairs_path)
scalar_slope_td <- fit_scalar_semivariance_slopes(scalar_pairs)
scalar_slope_td[, metric_clean := factor(
  clean_metric_label(metric_label),
  levels = rev(clean_metric_label(unique(scalar_slope_td[order(metric_order), metric_label])))
)]
scalar_slope_td[, treatment_label := factor(treatment_label, levels = c("Control", "Warmed"))]
scalar_slope_td[, depth_label := factor(depth_label, levels = depth_levels)]

scalar_slope_td[
  ,
  .(
    metric,
    treatment_label,
    depth_label,
    n_pairs,
    intercept,
    slope_per_log2_day_interval,
    r2
  )
]
```

### Two-Panel Slope Figure

```{r plot-slope-figure, fig.width=14, fig.height=6.6}
make_offset_plot_data <- function(dt, label_col, order_col, y_levels = NULL) {
  axis_dt <- unique(dt[, .SD, .SDcols = c(label_col, order_col)])
  setnames(axis_dt, c(label_col, order_col), c("label", "order"))
  axis_dt[, base_y := max(order, na.rm = TRUE) - order + 1]
  plot_dt <- merge(dt, axis_dt, by.x = label_col, by.y = "label", all.x = TRUE)
  plot_dt[, y_position := base_y + unname(depth_offsets[as.character(depth_label)])]
  list(plot_dt = plot_dt, axis_dt = axis_dt)
}

distance_axis <- data.table(
  distance_label = factor(distance_levels, levels = rev(distance_levels)),
  distance_order = seq_along(distance_levels)
)
distance_axis[, distance_order := max(distance_order) - distance_order + 1]
distance_slope_plot <- merge(distance_slope_td, distance_axis, by = "distance_label", all.x = TRUE)
distance_slope_plot[, y_position := distance_order + unname(depth_offsets[as.character(depth_label)])]
distance_segments <- dcast(
  distance_slope_plot,
  depth_label + distance_label + y_position ~ treatment_label,
  value.var = "slope_per_log2_day_interval"
)

p_distance_slope <- ggplot(distance_slope_plot, aes(
  x = slope_per_log2_day_interval,
  y = y_position,
  color = treatment_label,
  shape = depth_label
)) +
  geom_hline(
    data = distance_axis,
    aes(yintercept = distance_order),
    color = "#EEF0F3",
    linewidth = 0.4
  ) +
  geom_vline(xintercept = 0, color = "#6B7280", linewidth = 0.35) +
  geom_segment(
    data = distance_segments[is.finite(Control) & is.finite(Warmed)],
    aes(x = Control, xend = Warmed, y = y_position, yend = y_position),
    inherit.aes = FALSE,
    color = "#B8BDC7",
    linewidth = 0.45
  ) +
  geom_point(size = 3.0, stroke = 1.0, na.rm = TRUE) +
  scale_color_manual(values = treatment_cols) +
  scale_shape_manual(values = depth_shapes, breaks = names(depth_shapes), labels = depth_display_labels) +
  scale_x_continuous(
    labels = label_number(accuracy = 0.01),
    limits = c(-0.01, 0.18),
    breaks = seq(0, 0.18, by = 0.03),
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  scale_y_continuous(
    breaks = distance_axis[order(distance_order), distance_order],
    labels = as.character(distance_axis[order(distance_order), distance_label]),
    expand = expansion(add = 0.55)
  ) +
  guides(
    color = guide_legend(order = 1, override.aes = list(shape = 16)),
    shape = guide_legend(order = 2, override.aes = list(color = "#4B5563"))
  ) +
  labs(x = "Distance slope per log2(day interval)", y = NULL) +
  theme_trace() +
  theme(legend.position = "none", plot.margin = margin(5.5, 12, 5.5, 28))

metric_axis <- unique(scalar_slope_td[, .(metric_clean, metric_order)])
metric_axis[, base_y := max(metric_order, na.rm = TRUE) - metric_order + 1]
scalar_slope_plot <- merge(scalar_slope_td, metric_axis, by = "metric_clean", all.x = TRUE)
scalar_slope_plot[, y_position := base_y + unname(depth_offsets[as.character(depth_label)])]
scalar_segments <- dcast(
  scalar_slope_plot,
  depth_label + metric_clean + y_position ~ treatment_label,
  value.var = "slope_per_log2_day_interval"
)

p_scalar_slope <- ggplot(scalar_slope_plot, aes(
  x = slope_per_log2_day_interval,
  y = y_position,
  color = treatment_label,
  shape = depth_label
)) +
  geom_hline(
    data = metric_axis,
    aes(yintercept = base_y),
    color = "#EEF0F3",
    linewidth = 0.4
  ) +
  geom_vline(xintercept = 0, color = "#6B7280", linewidth = 0.35) +
  geom_segment(
    data = scalar_segments[is.finite(Control) & is.finite(Warmed)],
    aes(x = Control, xend = Warmed, y = y_position, yend = y_position),
    inherit.aes = FALSE,
    color = "#B8BDC7",
    linewidth = 0.45
  ) +
  geom_point(size = 2.7, stroke = 1.0, na.rm = TRUE) +
  scale_color_manual(values = treatment_cols) +
  scale_shape_manual(values = depth_shapes, breaks = names(depth_shapes), labels = depth_display_labels) +
  scale_x_continuous(
    labels = label_number(accuracy = 0.01),
    limits = c(-0.05, 0.40),
    breaks = seq(0, 0.40, by = 0.10),
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  scale_y_continuous(
    breaks = metric_axis[order(base_y), base_y],
    labels = as.character(metric_axis[order(base_y), metric_clean]),
    expand = expansion(add = 0.55)
  ) +
  guides(
    color = guide_legend(order = 1, override.aes = list(shape = 16)),
    shape = guide_legend(order = 2, override.aes = list(color = "#4B5563"))
  ) +
  labs(x = "Scaled semivariance per log2(day interval)", y = NULL) +
  theme_trace() +
  theme(legend.position = "none", plot.margin = margin(5.5, 5.5, 5.5, 56))

slope_legend <- get_legend(
  p_distance_slope + theme(legend.position = "top")
)

p_slope_two_panel <- plot_grid(
  slope_legend,
  plot_grid(
    p_distance_slope,
    p_scalar_slope,
    ncol = 2,
    align = "h",
    axis = "tb",
    labels = c("A", "B"),
    label_size = 12,
    label_fontface = "bold",
    rel_widths = c(0.92, 1.22)
  ),
  ncol = 1,
  rel_heights = c(0.08, 1)
)

p_slope_two_panel

if (write_outputs) {
  ggsave(
    file.path(fig_dir, "fticr_q2_temporal_slopes_two_panel_interactive.png"),
    p_slope_two_panel,
    width = 14,
    height = 6.6,
    units = "in",
    dpi = 320,
    bg = "white"
  )
  ggsave(
    file.path(fig_dir, "fticr_q2_temporal_slopes_two_panel_interactive.pdf"),
    p_slope_two_panel,
    width = 14,
    height = 6.6,
    units = "in",
    bg = "white"
  )
}
```

## Autocorrelation Terms With All Van Krevelen Classes

### Build Metric Table

This section expands the scalar metric table to include formula-count fractions for every van Krevelen class.

```{r build-vk-metrics}
if (use_cached_autocorrelation && file.exists(cached_autocorr_metric_path)) {
  metric_dt <- fread(cached_autocorr_metric_path)
  metric_order_dt <- unique(metric_dt[, .(metric, metric_label, metric_group, metric_order)])
  setorder(metric_order_dt, metric_order)
} else {
  long_dt <- fread(long_metric_path)
  long_dt[, date := as.IDate(date)]

  base_metric_map <- data.table(
    metric = c("richness", "mean_nosc", "mean_hc", "mean_oc", "mean_ai", "frac_n_containing"),
    metric_label = c(
      "Formula richness",
      "Mean NOSC",
      "Mean H/C",
      "Mean O/C",
      "Aromaticity",
      "N-containing formulae (%)"
    ),
    metric_group = c(rep("Bulk molecular metrics", 5), "Elemental formula fractions"),
    metric_order = seq_len(6L)
  )

  base_dt <- copy(long_dt[metric %in% base_metric_map$metric])
  base_dt[, c("metric_label", "metric_order") := NULL]
  base_dt <- merge(base_dt, base_metric_map, by = "metric", all.x = TRUE)

  mol <- fread(mol_path)
  formula_col <- "Molecular Formula"
  if ("Is Isotopologue" %in% names(mol)) {
    mol <- mol[safe_numeric(`Is Isotopologue`) == 0 | is.na(`Is Isotopologue`)]
  }
  for (nm in c("C", "H", "O", "N", "S", "P")) mol[, (nm) := safe_numeric(get(nm))]
  mol <- mol[!is.na(C) & !is.na(H) & C > 0]
  for (nm in c("O", "N", "S", "P")) mol[is.na(get(nm)), (nm) := 0]
  mol[, HC := H / C]
  mol[, OC := O / C]
  mol[, AI_mod_denominator := C - O - S - N - P]
  mol[, AI_mod := (1 + C - O - S - 0.5 * (H + N + P)) / AI_mod_denominator]
  mol[AI_mod_denominator <= 0 | !is.finite(AI_mod) | AI_mod < 0, AI_mod := 0]
  mol[, vk_class := classify_vk(HC, OC, AI_mod)]
  mol <- unique(mol, by = formula_col)

  intensity <- fread(intensity_path)
  setnames(intensity, formula_col, "formula")
  blank_cols <- grep("Blank", names(intensity), value = TRUE, ignore.case = TRUE)
  field_cols <- grep("^Sihi_60398_[0-9]+_r[0-9]+_", names(intensity), value = TRUE)
  blank_formulas <- character()
  if (length(blank_cols) > 0L) {
    blank_detected <- rowSums(!is.na(intensity[, ..blank_cols])) > 0
    blank_formulas <- intensity[blank_detected, formula]
  }

  valid_formula_set <- setdiff(mol[[formula_col]], blank_formulas)
  formula_class <- mol[get(formula_col) %in% valid_formula_set, .(formula = get(formula_col), vk_class)]
  setkey(formula_class, formula)

  sample_map <- data.table(sample_col = field_cols)
  sample_map[, sample_num := suppressWarnings(as.integer(sub("^Sihi_60398_([0-9]+)_r[0-9]+_.*$", "\\1", sample_col)))]
  sample_map <- sample_map[!is.na(sample_num)]

  vk_levels <- c("aromatic", "carbohydrate", "condensed_aromatic", "lignin_like", "lipid", "other")
  vk_metric_map <- data.table(
    vk_class = vk_levels,
    metric = paste0("vk_frac_", vk_levels),
    metric_label = c(
      "Aromatic formulae (%)",
      "Carbohydrate formulae (%)",
      "Condensed aromatic formulae (%)",
      "Lignin-like formulae (%)",
      "Lipid formulae (%)",
      "Other formulae (%)"
    ),
    metric_group = "Van Krevelen class fractions",
    metric_order = seq.int(max(base_metric_map$metric_order) + 1L, max(base_metric_map$metric_order) + length(vk_levels))
  )

  target_samples <- unique(long_dt[, .(sample_num, plot, date, treatment, treatment_label, depth_cm, depth_label)])
  vk_rows <- vector("list", nrow(target_samples))
  for (i in seq_len(nrow(target_samples))) {
    snum <- target_samples$sample_num[[i]]
    cols <- sample_map[sample_num == snum, sample_col]
    if (length(cols) == 0L) next
    detected <- rowSums(!is.na(intensity[, ..cols])) > 0
    detected_formulas <- intersect(intensity[detected, formula], valid_formula_set)
    if (length(detected_formulas) < 5L) next
    classes <- formula_class[J(detected_formulas), on = "formula", nomatch = 0, vk_class]
    class_counts <- table(factor(classes, levels = vk_levels))
    total <- sum(class_counts)
    if (!is.finite(total) || total == 0) next
    vk_rows[[i]] <- cbind(
      target_samples[i],
      data.table(vk_class = vk_levels, value = as.numeric(class_counts) / total, multiplier = 100)
    )
  }

  vk_dt <- rbindlist(vk_rows, fill = TRUE)
  vk_dt <- merge(vk_dt, vk_metric_map, by = "vk_class", all.x = TRUE)
  vk_dt[, value_plot := value * multiplier]
  vk_dt[, vk_class := NULL]

  metric_dt <- rbindlist(list(base_dt, vk_dt), fill = TRUE)
  metric_dt <- metric_dt[is.finite(value_plot)]
  metric_dt[, date := as.IDate(date)]
  metric_order_dt <- unique(metric_dt[, .(metric, metric_label, metric_group, metric_order)])
  setorder(metric_order_dt, metric_order)

  if (write_outputs) {
    fwrite(metric_dt, file.path(out_dir, "fticr_q2_autocorrelation_metric_plot_data_vk_classes_interactive.csv"))
  }
}

metric_order_dt
```

### Fit CAR(1) Autocorrelation Terms

```{r fit-car1-terms}
if (use_cached_autocorrelation && file.exists(cached_autocorr_path)) {
  autocorr_dt <- fread(cached_autocorr_path)
} else {
  autocorr_dt <- fit_car1_by_treatment_depth(metric_dt, metric_order_dt)
}

autocorr_dt[, metric_clean := factor(clean_metric_label(metric_label), levels = rev(metric_order_dt$metric_label))]
autocorr_dt[, treatment_label := factor(treatment_label, levels = c("Control", "Warmed"))]
autocorr_dt[, depth_label := factor(depth_label, levels = depth_levels)]

if (write_outputs) {
  fwrite(autocorr_dt, file.path(out_dir, "fticr_q2_autocorrelation_terms_by_treatment_depth_vk_classes_interactive.csv"))
}

autocorr_dt[
  ,
  .(
    metric,
    treatment_label,
    depth_label,
    n_samples,
    phi_per_30d,
    half_life_days,
    days_to_corr05,
    delta_aic_car1_minus_null
  )
]
```

### Autocorrelation Term Figure

```{r plot-autocorrelation-terms, fig.width=11, fig.height=7.8}
axis_dt <- unique(autocorr_dt[, .(metric_clean, metric_order)])
axis_dt[, base_y := max(metric_order, na.rm = TRUE) - metric_order + 1]
autocorr_plot_dt <- merge(autocorr_dt, axis_dt, by = "metric_clean", all.x = TRUE)
autocorr_plot_dt[, y_position := base_y + unname(depth_offsets[as.character(depth_label)])]
autocorr_segments <- dcast(
  autocorr_plot_dt,
  depth_label + metric_clean + y_position ~ treatment_label,
  value.var = "phi_per_30d"
)

p_autocorr <- ggplot(autocorr_plot_dt, aes(
  x = phi_per_30d,
  y = y_position,
  color = treatment_label,
  shape = depth_label
)) +
  geom_hline(
    data = axis_dt,
    aes(yintercept = base_y),
    color = "#EEF0F3",
    linewidth = 0.4
  ) +
  geom_vline(xintercept = 0, color = "#6B7280", linewidth = 0.35) +
  geom_segment(
    data = autocorr_segments[is.finite(Control) & is.finite(Warmed)],
    aes(x = Control, xend = Warmed, y = y_position, yend = y_position),
    inherit.aes = FALSE,
    color = "#B8BDC7",
    linewidth = 0.45
  ) +
  geom_point(size = 2.8, stroke = 1.0, na.rm = TRUE) +
  scale_color_manual(values = treatment_cols) +
  scale_shape_manual(values = depth_shapes, breaks = names(depth_shapes), labels = depth_display_labels) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.2),
    labels = label_number(accuracy = 0.1),
    expand = expansion(mult = c(0.01, 0.04))
  ) +
  scale_y_continuous(
    breaks = axis_dt[order(base_y), base_y],
    labels = as.character(axis_dt[order(base_y), metric_clean]),
    expand = expansion(add = 0.58)
  ) +
  guides(
    color = guide_legend(order = 1, override.aes = list(shape = 16)),
    shape = guide_legend(order = 2, override.aes = list(color = "#4B5563"))
  ) +
  labs(x = "30-day temporal autocorrelation", y = NULL) +
  theme_trace()

p_autocorr

if (write_outputs) {
  ggsave(
    file.path(fig_dir, "fticr_q2_autocorrelation_terms_vk_classes_interactive.png"),
    p_autocorr,
    width = 11,
    height = 7.8,
    units = "in",
    dpi = 320,
    bg = "white"
  )
  ggsave(
    file.path(fig_dir, "fticr_q2_autocorrelation_terms_vk_classes_interactive.pdf"),
    p_autocorr,
    width = 11,
    height = 7.8,
    units = "in",
    bg = "white"
  )
}
```

## Optional: Autocorrelation Decay Curve

Edit `selected_metric`, `selected_treatment`, and `selected_depth` in the setup chunk, then rerun this section.

```{r selected-car1-decay, fig.width=7.5, fig.height=4.5}
selected_depth_raw <- names(depth_display_labels)[match(selected_depth, depth_display_labels)]
selected_row <- autocorr_dt[
  metric == selected_metric &
    treatment_label == selected_treatment &
    depth_label == selected_depth_raw
]

if (nrow(selected_row) != 1L || !is.finite(selected_row$phi_per_30d)) {
  message("No finite CAR(1) estimate found for this selected metric/treatment/depth.")
} else {
  lag_grid <- data.table(lag_days = seq(0, 365, by = 7))
  lag_grid[, autocorrelation := selected_row$phi_per_30d^(lag_days / 30)]
  ggplot(lag_grid, aes(x = lag_days, y = autocorrelation)) +
    geom_hline(yintercept = 0.05, color = "#6B7280", linewidth = 0.35, linetype = "dashed") +
    geom_line(color = "#2B6CB0", linewidth = 0.9) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
    scale_x_continuous(breaks = c(0, 30, 60, 90, 180, 365)) +
    labs(
      x = "Lag (days)",
      y = "Implied autocorrelation"
    ) +
    theme_trace()
}

selected_row[
  ,
  .(
    metric,
    treatment_label,
    depth_label,
    phi_per_30d,
    half_life_days,
    days_to_corr05,
    delta_aic_car1_minus_null
  )
]
```

## Takeaways

Use the slope figure for the question:

```text
Does dissimilarity increase as the time between samples doubles?
```

Use the CAR(1) figure for the question:

```text
How much temporal memory remains at a standardized 30-day lag?
```

These are related but not identical. Semivariance can increase steeply with lag even when 30-day autocorrelation is high, because strong short-term memory can still decay substantially over longer intervals.
